The Desert of the Reals: Floating-Point Arithmetic on Deterministic Systems

### 2024-04-10 mss.tex
Fix bib

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              Manuscript submitted for review.\\
              ~ \\
              Address author correspondence to \patp{lagrev-nocfep}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Floating-Point Arithmetic on Deterministic Systems}
\fancyfoot[LE,RO]{\thepage}

@@

@@

Early numeric computing tended to focus on problems of interest to military and national security applications, such as the solution of differential equations and numerical optimization.  Such calculations typically involve arrays, and linear algebra was elaborated hand-in-hand with digital computing techniques in software and hardware.  Numerics assumed prominence for a wider audience with the rise of gaming on personal computers, although these algorithms emphasized speed over exactness.\footnote{This is reflected in algorithms such as the “fast inverse square root”, which permits a degree of inaccuracy in exchange for a substantial speedup.}  To this point in the history of computing, most software either ran on a single platform for its lifetime (as with supercomputing) or did not require portably deterministic algorithms (as with gaming).\footnote{To be clear, 3D gaming algorithms are deterministic (assuming no random sources are used), but exact error values are often not reproducible across platforms, nor was such portability a design criterion.  The reasons for this are discussed below.}

@@

In the historical development of mathematics, logical problems in each set of numbers drove the discovery and elaboration of more elaborate algebras.  For instance, in the field of natural numbers $\mathbb{N}$, the operation of addition $+$ or multiplication $\times$ produces a value within the set; however, permitting subtraction $-$ of a larger number from a smaller number can result in a value inexpressible in $\mathbb{N}$.  This motivated the introduction of the integers $\mathbb{Z}$, augmenting the numbers from zero to (positive) infinity with the negative numbers.  Division $/$ similarly produced a crisis when applied to values which did not have a whole-number ratio between them, a situation resolved by the Pythagorean\footnote{A legendary attribution, alas, predicated on the Pythagorean discovery of the irrational numbers as a separate class thereby inducing a crisis.  https://plato.stanford.edu/entries/pythagoras/} innovation of the rational numbers or fractions as a class $\mathbb{Q}$.  Ultimately, the common reference set for engineering mathematics (and the human understanding of the continuum such as measurement) is the set of real numbers.  The set of real numbers, denoted by $\mathbb{R}$, is characterized by its continuity, implying that for any two distinct values within this set, there exists a difference, no matter how small.

\sloppy
Since the operations and conventions of $\mathbb{R}$ have been found to be so useful, it is desirable to extend the semantics to computer programming.  However, digital computers, by virtue of their binary representation, effectively use natural numbers $\mathbb{N}$ to represent numbers (to the limit of memory rather than positive infinity $+\infty$).  Several schemes permit a computer integer to be interpreted as if it were a number with a fractional part, including a scaling factor, fixed-point representation,\footnote{Hand-in-hand with the development of linear algebra, machines such as {\sc eniac} and {\sc maniac} were employed in the 1940s for solving thermonuclear reaction calculations and neutron diffusion equations.  Under the direction of John von Neumann, it appears that some calculations did experimentally involve a floating-point scheme, although this was later rejected definitively in favor of fixed-point arithmetic.  See \citet{Kahan1997a}, p.~3 on floating-point representations.}

@@

\subsection{{\sc ieee} 754 Basics}

@@

\begin{lstlisting}[style=listingblock]
SEEE.EEEE.EFFF.FFFF.FFFF.FFFF.FFFF.FFFF
\end{lstlisting}
\noindent
where \texttt{S} is the sign bit, \texttt{0} for positive ($+$) and \texttt{1} for negative ($-$); \texttt{E} is the exponent in base-2 (8 bits); and \texttt{F} is the significand (23 bits).  The exponent is actually calculated at an offset bias of 127 ($2^{7}$) so that a more expressive range of orders of magnitude can be covered.  The significand has an implied leading \texttt{1} bit unless all are zero.  To wit,

@@

Since the {\sc ieee} 754 floating-point format packs values of different kind together bitwise, conventional integer operations such as left shift \texttt{<<} and addition \texttt{+} do not trivially apply.\footnote{For instance, left-shifting \texttt{1.0} does not result in \texttt{2.0} but \texttt{1.7014118e38}; we leave the mathematics of why as an exercise to the reader.}

@@

For most values of the exponent \texttt{E}, the difference between two discrete values is determined by the absolute magnitude of the significand \texttt{S}.  The difference between serial values is
\begin{align}
  \Delta\texttt{S} &= \texttt{000.0000.0000.0000.0000.0001}_{2} \nonumber \\
  & = 1.00000011920928955078125 - 1.0 \nonumber \\
  & = 0.00000011920928955078125 \nonumber \\
  & = 2^{-23}\textrm{.}
\end{align}

@@

\lstset{language=Python}
\begin{lstlisting}[style=listingblock_python]
In [1]: (1.1-0.3)-0.8
Out[1]: 0.0

@@

For instance, {\sc ieee} 754 defines a range of numbers as “not-a-number” values, or NaNs.  Per the specification, a NaN can be a signalling NaN, meaning that it intends to flag and possibly disrupt a problematic computation;\footnote{“C support for signaling NaNs, or for auxiliary information that could be encoded in NaNs, is problematic.  Trap handling varies widely among implementations. Implementation mechanisms may trigger signaling NaNs, or fail to, in mysterious ways.  The {\sc iec} 60559 floating-point standard recommends that NaNs propagate; but it does not require this and not all implementations do.”  \citetext{\citet{Jones2008}, p.~339}} or a quiet NaN, which does not raise such an exception and merely yields a result with the NaN propagated to the final result.\footnote{In Urbit, the Vere runtime unifies NaNs, meaning that any bitwise information which may be encoded in the significand field—the “NaN payload”—is thrown away.}  Not all processors implement this part of {\sc ieee} 754 correctly:  “The Motorola {\sc dsp563ccc} does not support NaNs or infinities.  Floating-point arithmetic operations do not overflow to infinity; they saturate at the maximum representable value” \citetext{\citet{Jones2008}, p.~338}.

@@

\section{Urbit's Implementation of {\sc ieee} 754}

@@

%\lstset{language=Hoon}
\begin{lstlisting}[style=listingblock]
++  add
  |=  [a=[e=@s a=@u] b=[e=@s a=@u] e=?]  ^-  fn
  =+  q=(dif:si e.a e.b)
  |-  ?.  (syn:si q)  $(b a, a b, q +(q))
  ?:  e
    [%f & e.b (^add (lsh [0 (abs:si q)] a.a) a.b)]
  =+  [ma=(met 0 a.a) mb=(met 0 a.b)]
  =+  ^=  w  %+  dif:si  e.a  %-  sun:si
    ?:  (gth prc ma)  (^sub prc ma)  0
  =+  ^=  x  %+  sum:si  e.b  (sun:si mb)
  ?:  =((cmp:si w x) --1)
    ?-  r
      %z  (lug %fl a &)  %d  (lug %fl a &)
      %a  (lug %lg a &)  %u  (lug %lg a &)
      %n  (lug %na a &)
    ==
  (rou [e.b (^add (lsh [0 (abs:si q)] a.a) a.b)])
\end{lstlisting}

@@

\lstset{language=C}
\begin{lstlisting}[style=listingblock_c]
u3_noun u3qet_add(u3_atom a, u3_atom b, u3_atom r) {
  union sing c, d, e;
  // set IEEE 754 rounding mode
  _set_rounding(r);
  // unwrap nouns into C-typed values
  c.c = u3r_word(0, a);
  d.c = u3r_word(0, b);
  // perform addition and unify NaN
  e.s = _nan_unify(f32_add(c.s, d.s));

@@

\section{Deterministic Computation with a Fractional Part}

@@

\begin{lstlisting}[style=listingblock_c]
#include <stdio.h>

@@

In 2015, John Gustafson proposed a new standard for representing values drawn from $\mathbb{R}$ called universal numbers or unums \citep{Gustafson2015}.\footnote{See also \citet{Gustafson2017}, \citet{Gustafson2017a}, and \citet{Gustafson2017b}.}  The current version of unums, called posits, supports interval arithmetic and greater resolution near 1.0, at the cost of decreased resolution for extremely large and extremely small values.  Unums also guarantee associativity and distributivity of operations.

Gustafson's criticisms of {\sc ieee} 754 focused on determinism and exactness; underflow and overflow; fixed bit widths for mantissa and exponent; rounding; and the large wasted block of NaNs \citep{Risse2016}.  Unums likewise must provide sign, exponent with bias, and significand; they additionally signal whether the value is an interval.  Unlike {\sc ieee} 754's use of multiple bit widths, 32-bit posits are argued to be sufficient for almost all applications.

The unum proposal appears to have settled somewhat after its initial state of relative flux (as ``posits'' or Type III unums, cf. \citet{Posit2022}).  Although most implementations have been in software, the project has been specified in Verilog several times and implemented on {\sc fpga}s (VividSparks, \url{https://vividsparks.tech/}, and \citet{Chen2018}).  Intriguingly, some initial work has been carried out towards a fundamental {\sc blas}-like library built on posits \citep{vanDam2019}.

@@

Several recent efforts on Urbit have encountered difficulties in producing reliably deterministic and sufficiently fast floating-point calculations on a Nock-based system.\footnote{Cf. UrWasm, which as a WebAssembly implementation directly relies on floating-point computations \citep{UrWasm} (see e.g. \citet{KloudKoder2022} for a discussion of the issues involved, and in particular \citet{Afonin2024} in the current text), \href{https://github.com/Quodss/urwasm}{\texttt{Quodss/urwasm}}; \texttt{/lib/math}, implemented purely in Hoon, \href{https://github.com/sigilante/libmath}{\texttt{sigilante/libmath}}; and Lagoon, which includes {\sc ieee} 754 operations with its linear algebra library, \href{https://github.com/urbit/numerics}{\texttt{urbit/numerics}} pending development and integration into Urbit.}  We anticipate that, water finding its own level, each will adopt a suitable deterministic solution for evaluation in Nock.  We do not anticipate these to be the last foundational numerical libraries built on Urbit, but instead among the first.  Thus we have documented the paths we have explored as an annotated map for future travelers in search of a one true representation for continuous mathematics. \tombstone{}

% \selectlanguage{USenglish}
\printbibliography
\end{document}

>>>>

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:1 (2024):  93–131. \\
              ~ \\
              Address author correspondence to \patp{lagrev-nocfep}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Floating-Point Arithmetic on Deterministic Systems}
\fancyfoot[LE,RO]{\thepage}

@@

\pagenumbering{arabic}
\setcounter{page}{93}

@@

\sloppy
Early numeric computing tended to focus on problems of interest to military and national security applications, such as the solution of differential equations and numerical optimization.  Such calculations typically involve arrays, and linear algebra was elaborated hand-in-hand with digital computing techniques in software and hardware.  Numerics assumed prominence for a wider audience with the rise of gaming on personal computers, although these algorithms emphasized speed over exactness.\footnote{This is reflected in algorithms such as the “fast inverse square root”, which permits a degree of inaccuracy in exchange for a substantial speedup.}  To this point in the history of computing, most software either ran on a single platform for its lifetime (as with supercomputing) or did not require portably deterministic algorithms (as with gaming).\footnote{To be clear, 3D gaming algorithms are deterministic (assuming no random sources are used), but exact error values are often not reproducible across platforms, nor was such portability a design criterion.  The reasons for this are discussed below.}

@@

In the historical development of mathematics, logical problems in each set of numbers drove the discovery and elaboration of more elaborate algebras.  For instance, in the field of natural numbers $\mathbb{N}$, the operation of addition $+$ or multiplication $\times$ produces a value within the set; however, permitting subtraction $-$ of a larger number from a smaller number can result in a value inexpressible in $\mathbb{N}$.  This motivated the introduction of the integers $\mathbb{Z}$, augmenting the numbers from zero to (positive) infinity with the negative numbers.  Division $/$ similarly produced a crisis when applied to values which did not have a whole-number ratio between them, a situation resolved by the Pythagorean\footnote{A legendary attribution, alas, predicated on the Pythagorean discovery of the irrational numbers as a separate class thereby inducing a crisis \citep{SEPPythagoras}.} innovation of the rational numbers or fractions as a class $\mathbb{Q}$.  Ultimately, the common reference set for engineering mathematics (and the human understanding of the continuum such as measurement) is the set of real numbers.  The set of real numbers, denoted by $\mathbb{R}$, is characterized by its continuity, implying that for any two distinct values within this set, there exists a difference, no matter how small.

\sloppy
Since the operations and conventions of $\mathbb{R}$ have been found to be so useful, it is desirable to extend the semantics to computer programming.  However, digital computers, by virtue of their binary representation, effectively use natural numbers $\mathbb{N}$ to represent numbers (to the limit of memory rather than positive infinity $+\infty$).  Several schemes permit a computer integer to be interpreted as if it were a number with a fractional part, including a scaling factor, fixed-point representation,\footnote{Hand-in-hand with the development of linear algebra, machines such as {\sc eniac} and {\sc maniac} were employed in the 1940s for solving thermonuclear reaction calculations and neutron diffusion equations.  Under the direction of John von Neumann, it appears that some calculations did experimentally involve a floating-point scheme, although this was later rejected definitively in favor of fixed-point arithmetic.  See \citet{Kahan1997a}, p.~3, on floating-point representations.}

@@

\subsection[\textsc{ieee} 754 Basics]{IEEE 754 Basics}

@@

\begin{lstlisting}[style=listingcode]
SEEE.EEEE.EFFF.FFFF.FFFF.FFFF.FFFF.FFFF
\end{lstlisting}
\noindent
where \texttt{S} is the sign bit, \texttt{0} for positive ($+$) and \texttt{1} for negative ($-$); \texttt{E} is the exponent in base-2 (8 bits); and \texttt{F} is the significand (23 bits).  The exponent is actually calculated at an offset bias of 127 ($2^{7}$) so that a more expressive range of orders of magnitude can be covered.  The significand has an implied leading \texttt{1} bit unless all are zero.  To wit,

@@

Since the {\sc ieee} 754 floating-point format packs values of different kind together bitwise, conventional integer operations such as left shift $<<$ and addition $+$ do not trivially apply.\footnote{For instance, left-shifting \texttt{1.0} does not result in \texttt{2.0} but \texttt{1.7014118e38}; we leave the mathematics of why as an exercise to the reader.}

@@

For most values of the exponent \texttt{E}, the difference between two discrete values is determined by the absolute magnitude of the significand \texttt{S}.  The difference between serial values is
\begin{align}
  \Delta\texttt{S} &= \texttt{␣000.0000.0000.0000.0000.0001}_{2} \nonumber \\
  & = 1.00000011920928955078125 - 1.0 \nonumber \\
  & = 0.00000011920928955078125 \nonumber \\
  & = 2^{-23}\textrm{.}
\end{align}

@@

\lstset{language=Python}
\begin{lstlisting}[style=listingcode_python]
In [1]: (1.1-0.3)-0.8
Out[1]: 0.0

@@

For instance, {\sc ieee} 754 defines a range of numbers as “not-a-number” values, or NaNs.  Per the specification, a NaN can be a signalling NaN, meaning that it intends to flag and possibly disrupt a problematic computation;\footnote{“C support for signaling NaNs, or for auxiliary information that could be encoded in NaNs, is problematic.  Trap handling varies widely among implementations. Implementation mechanisms may trigger signaling NaNs, or fail to, in mysterious ways.  The {\sc iec} 60559 floating-point standard recommends that NaNs propagate; but it does not require this and not all implementations do.”  \citetext{\citet{Jones2008}, p.~339}} or a quiet NaN, which does not raise such an exception and merely yields a result with the NaN propagated to the final result.\footnote{In Urbit, the Vere runtime unifies NaNs, meaning that any bitwise information which may be encoded in the significand field—the “NaN payload”—is thrown away.  This is called the ``canonical NaN''.}  Not all processors implement this part of {\sc ieee} 754 correctly:  “The Motorola {\sc dsp563ccc} does not support NaNs or infinities.  Floating-point arithmetic operations do not overflow to infinity; they saturate at the maximum representable value” \citetext{\citet{Jones2008}, p.~338}.

@@

\section{Urbit's Implementation of \textsc{\textbf{ieee}} 754}

@@

%\lstset{language=Hoon}
\begin{lstlisting}[style=listingcode]
++  add
  |=  [a=[e=@s a=@u] b=[e=@s a=@u] e=?]  ^-  fn
  =+  q=(dif:si e.a e.b)
  |-  ?.  (syn:si q)  $(b a, a b, q +(q))
  ?:  e
    [%f & e.b (^add (lsh [0 (abs:si q)] a.a) a.b)]
  =+  [ma=(met 0 a.a) mb=(met 0 a.b)]
  =+  ^=  w  %+  dif:si  e.a  %-  sun:si
    ?:  (gth prc ma)  (^sub prc ma)  0
  =+  ^=  x  %+  sum:si  e.b  (sun:si mb)
  ?:  =((cmp:si w x) --1)
    ?-  r
      %z  (lug %fl a &)  %d  (lug %fl a &)
      %a  (lug %lg a &)  %u  (lug %lg a &)
      %n  (lug %na a &)
    ==
  (rou [e.b (^add (lsh [0 (abs:si q)] a.a) a.b)])
\end{lstlisting}

@@

\lstset{language=C}
\begin{lstlisting}[style=listingcode_c]
u3_noun u3qet_add(u3_atom a, u3_atom b, u3_atom r) {
  union sing c, d, e;
  // set IEEE 754 rounding mode
  _set_rounding(r);
  // unwrap nouns into C-typed values
  c.c = u3r_word(0, a);
  d.c = u3r_word(0, b);
  // perform addition and unify NaN
  e.s = _nan_unify(f32_add(c.s, d.s));

@@

\section[Deterministic\,Computation\,with\,Fractional\,Part]{Deterministic Computation with a \\ Fractional Part}

@@

\begin{lstlisting}[style=listingcode_c]
#include <stdio.h>

@@

In 2015, John Gustafson proposed a new standard for representing values drawn from $\mathbb{R}$ called universal numbers or unums \citep{Gustafson2015}.\footnote{See also \citet{Gustafson2017}, \citet{Gustafson2017a}, and \citet{Gustafson2017b}.}  The current version supports interval arithmetic and greater resolution near 1.0, at the cost of decreased resolution for extremely large and extremely small values.  Unums also guarantee associativity and distributivity of operations.

Gustafson's criticisms of {\sc ieee} 754 focused on determinism and exactness; underflow and overflow; fixed bit widths for mantissa and exponent; rounding; and the large wasted block of NaNs \citep{Risse2016}.  Unums likewise must provide sign, exponent with bias, and significand; they may additionally signal whether the value is an interval.  Unlike {\sc ieee} 754's use of multiple bit widths, 32-bit ``posits'' (fixed-size unums intended to facilitate hardware requirements) are argued to be sufficient for almost all applications.

The unum proposal appears to have settled somewhat after its initial state of relative flux (as Type III unums, cf. \citet{Posit2022}).  Although most implementations have been in software, the project has been specified in Verilog several times and implemented on {\sc fpga}s (VividSparks, \url{https://vividsparks.tech/}, and \citet{Chen2018}).  Intriguingly, some initial work has been carried out towards a fundamental {\sc blas}-like library built on posits \citep{vanDam2019}.

@@

Several recent efforts on Urbit have encountered difficulties in producing reliably deterministic and sufficiently fast floating-point calculations on a Nock-based system.\footnote{Cf. UrWasm, which as a WebAssembly implementation directly relies on floating-point computations \citep{UrWasm} (see e.g. \citet{KloudKoder2022} for a discussion of the issues involved, and in particular \citeauthor{Afonin2024}, pp. 133–150 in this issue), \href{https://github.com/Quodss/urwasm}{\texttt{Quodss/urwasm}}; \texttt{/lib/math}, implemented purely in Hoon, \href{https://github.com/sigilante/libmath}{\texttt{sigilante/libmath}}; and Lagoon, which includes {\sc ieee} 754 operations with its linear algebra library, \href{https://github.com/urbit/numerics}{\texttt{urbit/numerics}} pending development and integration into Urbit.}  We anticipate that, water finding its own level, each will adopt a suitable deterministic solution for evaluation in Nock.  We do not anticipate these to be the last foundational numerical libraries built on Urbit, but instead among the first.  Thus we have documented the paths we have explored as an annotated map for future travelers in search of a one true representation for continuous mathematics. \tombstone{}

\selectlanguage{USenglish}
\printbibliography
\end{document}
